---
title: "4_GAMAnalysis"
output: html_document
date: "2025-08-09"
---

###Manuscript under review

**Title:** Mixed landscapes, social vulnerability, and climate change shape West Nile virus risk across the United States

**corresponding author:** Samantha Sambado (ssambado at stanford.ed)

##Purpose
Run GAM analysis

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

## upload necessary packages
library(ggcorrplot)
library(mgcv)
library(gam.hp)
library(gratia)

## upload necessary data
covariates <- read.csv("data/processed/gam_covariates_noR0.csv") 
rR0_peakmonth_county <- read.csv("data/processed/rR0_peakmonth_county_ssp245_hist.csv") 
```

##Step 1. Format GAM data
```{r}
## make sure GEOID is correct format
covariates_clean <- covariates %>% 
  dplyr::select(-X)  %>% 
  mutate(GEOID = str_pad(GEOID, width = 5, pad = 0), # make sure FIPS are 5 characters
         GEOID = as.character(GEOID)) 


rR0_peakmonth_county_clean <- rR0_peakmonth_county %>% dplyr::select(-X)  %>% 
  mutate(GEOID = str_pad(GEOID, width = 5, pad = 0), # make sure FIPS are 5 characters
         GEOID = as.character(GEOID)) 

## merge
gam_covariates_full <- covariates_clean %>% 
  left_join(rR0_peakmonth_county_clean, by = "GEOID")

write.csv(gam_covariates_full, file = "data/processed/gam_covariates_full.csv")

```

##Step 2. data checks
```{r}
## A. distribution of incidence
hist(gam_covariates_full$raw_incidence) # right skewed
hist(gam_covariates_full$log_incidence) # seems normal enough


## B. correlation of covariates
gam_covariates_slim <- gam_covariates_full %>% 
  dplyr::select(raw_incidence,
                raw_perc_over65,
                raw_log_edgedensity_mk2,
                raw_bls_agemploy,
                raw_svi_mean,
                raw_birds_tot,
                raw_peakmonths_rR0_mean)

cor_matrix <- cor(gam_covariates_slim, use = "complete.obs")

ggcorrplot(cor_matrix, 
           method = "square",     
           type = "upper",       
           lab = TRUE,        
           lab_size = 6, 
           show.legend =FALSE,
           tl.col = "black",        
           tl.srt = 45,)

## really the only concerning one is rR0 and SVI but probably heat islands/high heat is part of the SVI metric
```


##Step 3. Model Construction
```{r}
## build model based on prior attempts
mod_best_scaled <- gam(log_incidence ~
                         #climate 
                         s(scale_peakmonths_rR0_mean) +
                         
                         # landscape
                         s(scale_log_edgedensity_mk2) +
                         
                         # human
                         scale_bls_agemploy +
                         s(scale_svi_mean) + 
                         
                         # bird
                         scale_birds_tot +
                         
                         # controls
                         scale_perc_over65 +
                         s(lon, lat, bs = "gp"), 
                       
                       data = gam_covariates_full,
                       method = "REML")
```

##Step 4. Model diagnostics
```{r}
# check model summary
summary(mod_best_scaled)

# check residuals
appraise(mod_best_scaled)

# visualize smooth terms
draw(mod_best_scaled)
```
##Step 5. Calculate % of deviance explained for each predictor
```{r}
gam.hp_result <- gam.hp(mod_best_scaled)
gam.hp_result_table <- as.data.frame(gam.hp_result$hierarchical.partitioning)

names(gam.hp_result_table)[4] <- "percent"

gam.hp_result_table %>% 
  arrange(desc(percent))
```




